---
title: "marlatt1975"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

The purpose of this file is to simulate the data from Marlatt, Kosturn, & Lang (1975; https://doi.org/10.1037/0021-843X.84.6.652). We don't have their actual data file, but they displayed the means and standard deviations for their 6 groups in Figure 1 (p. 656). By design, there were 10 participants in each group.

Load the **tidyverse**.

```{r, warning = F, message = F}
library(tidyverse)
```

## Set up

In Table 1 (p. 656), the authors listed the means and standard deviations for the six groups on the primary outcome variable, the ounces of wine consumed in the taste-rating task. Here are those values.

```{r}
# save the values
m <- c(7.31, 4.86, 6.69, 6.90, 4.37, 4.66)
s <- c(2.93, 1.97, 3.56, 3.04, 2.38, 2.71)

# display them in a tibble
tibble(sex       = rep(c("m", "w"), each = 3),
       condition = rep(1:3, times = 2),
       mean      = m,
       sd        = s)
```

The ammount of ounces one drinks is a continuous variable with a lower limit of 0 (i.e., one cannot drink -2 ounces). Without further information, the gamma distribution seems like a good candedate to simulate such data. A difficulty is the `rgamma()` function is not parameterized in terms of the mean and standard deviation.

```{r}
rgamma
```

Rather, one can define gamma draws by wither the combination of `shape` and `rate` or of `shape` and `scale`. In his [text](https://sites.google.com/site/doingbayesiandataanalysis/), Kruschke provided a convenience function that takes means and standard deviations as input and returns the `shape` and `rate` values for the corresponding gamma distribution. Here we define that function.

```{r}
gammaShRaFromMeanSD = function(mean, sd) {
  if (mean <= 0) stop("mean must be > 0")
  if (sd   <= 0) stop("sd must be > 0")
  shape <- mean^2 / sd^2
  rate  <- mean   / sd^2
  return(list(shape = shape, rate = rate))
}
```

Now use Kruschke's `gammaShRaFromMeanSD()` function to compute the `shape` and `rate` parameters from the vectors of means (`m`) and standard deviations (`s`) from above.

```{r}
# compute and save the parameter values for men
m1 <- gammaShRaFromMeanSD(mean = m[1], sd = s[1])
m2 <- gammaShRaFromMeanSD(mean = m[2], sd = s[2])
m3 <- gammaShRaFromMeanSD(mean = m[3], sd = s[3])

# compute and save the parameter values for women
w1 <- gammaShRaFromMeanSD(mean = m[4], sd = s[4])
w2 <- gammaShRaFromMeanSD(mean = m[5], sd = s[5])
w3 <- gammaShRaFromMeanSD(mean = m[6], sd = s[6])
```

Here's a quick example of what we have saved.

```{r}
m1
```

To get a sense of how this all works, here's what the corresponding 6 gamma distributions would look like based on 10,000 simulation draws, each.

```{r}
# how many draws?
n <- 1e4

set.seed(1)

tibble(m.1 = rgamma(n = n, shape = m1$shape, rate = m1$rate),
       m.2 = rgamma(n = n, shape = m2$shape, rate = m2$rate),
       m.3 = rgamma(n = n, shape = m3$shape, rate = m3$rate),
       w.1 = rgamma(n = n, shape = w1$shape, rate = w1$rate),
       w.2 = rgamma(n = n, shape = w2$shape, rate = w2$rate),
       w.3 = rgamma(n = n, shape = w3$shape, rate = w3$rate)) %>% 
  pivot_longer(everything()) %>% 
  separate(name, into = c("sex", "condition")) %>% 
  
  # plot
  ggplot(aes(x = value)) +
  geom_histogram(binwidth = 1) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_cartesian(xlim = c(0, 20)) +
  facet_grid(sex ~ condition, labeller = label_both)
```

Now simulate the actual data, which we'll save as `marlatt1975`.

```{r}
# how many per group?
n <- 10

# I arrived at this seed value by a little trial and error
set.seed(3)

marlatt1975 <- tibble(
  m.1 = rgamma(n = n, shape = m1$shape, rate = m1$rate),
  m.2 = rgamma(n = n, shape = m2$shape, rate = m2$rate),
  m.3 = rgamma(n = n, shape = m3$shape, rate = m3$rate),
  w.1 = rgamma(n = n, shape = w1$shape, rate = w1$rate),
  w.2 = rgamma(n = n, shape = w2$shape, rate = w2$rate),
  w.3 = rgamma(n = n, shape = w3$shape, rate = w3$rate)) %>% 
  pivot_longer(everything(), values_to = "ounces") %>% 
  separate(name, into = c("sex", "condition")) %>% 
  mutate(condition = factor(condition,
                            labels = c("insulted only",
                                       "insulted with retalliation",
                                       "no-insult control"))) %>% 
  # reduce the precision of the values
  mutate(ounces = round(ounces, digits = 2))

# what have we done?
glimpse(marlatt1975)
```

Here's a plot of the `marlatt1975` data.

```{r}
marlatt1975 %>% 
  ggplot(aes(x = ounces)) +
  geom_histogram(binwidth = 1) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_cartesian(xlim = c(0, 20)) +
  facet_grid(sex ~ condition)
```

Here are the sample statistics.

```{r}
marlatt1975 %>% 
  group_by(sex, condition) %>% 
  summarise(m = mean(ounces) %>% round(digits = 2),
            s = sd(ounces) %>% round(digits = 2),
            n = n())
```

## Save

Now save the results in an externa `.rda` file.

```{r}
save(marlatt1975, file = "/Users/solomonkurz/Dropbox/Experimental-design-and-the-GLMM/sketches/data/marlatt1975.rda")
```

## Session information

```{r}
sessionInfo()
```

