---
title: "mason2014"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

The purpose of this file is to simulate the data from Mason et al (2014; https://doi.org/10.1016/j.rasd.2013.12.014). We don't have their actual data file, they displayed their data in Figure 2 (p. 340). Here we'll work through that figure to make a data file that approximates their original data.

Load the **tidyverse**.

```{r, warning = F, message = F}
library(tidyverse)
```

First make the person-specific data frames.

```{r}
# Sam
sam <- tibble(
  id      = "Sam",
  session = c(3, 4, 5, 6, 7, 8,
              9, 10, 11, 12, 15, 16, 19, 21, 22, 25, 32, 36, 42),
  count   = c(4, 13, 4, 6, 6, 6,
              38, 20, 34, 41, 17, 16, 38, 21, 28, 35, 56, 21, 40),
  phase   = rep(LETTERS[1:2], times = c(6, 13)))

# Ed
ed <- tibble(
  id      = "Ed",
  session = c(1, 2, 3, 4, 5, 6, 7, 9, 11,
              15, 16, 19, 22, 24, 25, 29, 30, 32, 34, 36, 37, 39, 42, 45),
  count   = c(5, 8, 2, 5, 1, 4, 11, 5, 2,
              27, 33, 29, 30, 33, 24, 24, 25, 30, 41, 38, 24, 21, 39, 27),
  phase   = rep(LETTERS[1:2], times = c(9, 15)))

# Brian
brian <- tibble(
  id      = "Brian",
  session = c(2:7, 16, 19, 22,
              25, 27, 29, 30:32, 33, 34, 42, 46, 47),
  count   = c(8, 1, 2, 6, 5, 1, 1, 3, 3,
              22, 27, 22, 15, 24, 11, 51, 6, 29, 20, 33),
  phase   = rep(LETTERS[1:2], times = c(9, 11)))
```

Now combine them and wrangle a little to make the `mason2014` data frame.

```{r}
mason2014 <- bind_rows(
  sam, ed, brian
) %>% 
  mutate(id = fct_rev(id),
         phase = factor(phase),
         session0 = session - 1) %>% 
  select(id, session, session0, phase, count)

# what have we done?
glimpse(mason2014)
head(mason2014)
```

We can look at the data with a version of Figure 2 from the original article.

```{r}
lines <- mason2014 %>% 
  distinct(id) %>% 
  mutate(xintercept = c(8.5, 13, 23.5))

mason2014 %>% 
  ggplot(aes(x = session, y = count, color = phase)) +
  geom_vline(data = lines,
             aes(xintercept = xintercept),
             color = "white") +
  geom_point(size = 2) +
  geom_line() +
  ylim(0, 60) +
  facet_wrap(~ id, ncol = 1) +
  theme(panel.grid = element_blank())
```

Now save the results in an external file.

```{r}
save(mason2014, file = "/Users/solomonkurz/Dropbox/Experimental-design-and-the-GLMM/sketches/data/mason2014.rda")
```

## Session information

```{r}
sessionInfo()
```

