---
title: "Post-test only designs with a large number of conditions: Milkman et al (2022)"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r, echo = F}
knitr::opts_chunk$set(fig.retina = 2.5)
```

Using a post-test only design, Milkman et al (2022) randomly assigned $689{,}693$ Walmart pharmacy patients into one of 23 conditions. Participants in 22 of the conditions received one of the 22 text-message reminders designed to encourage Walmart flu vaccination. Participants in the business-as-usual control condition received no such messages. The primary outcome was the proportion of persons in each condition who received the flu vaccine at Walmart by December 31st, 2020.

Milkman et al (2022) kindly made a subset of their data public on the OSF at https://osf.io/rn8tw/?view_only=546ed2d8473f4978b95948a52712a3c5. The purpose of this file is to show how to analyze their main results, similar to how they were presented in Figure 1 (p. 3).

Load the data and the primary **R** packages.

```{r, warning = F, message = F}
# packages
library(tidyverse)
library(brms)

# load the data
d <- readxl::read_excel("data/aggregate_data.xlsx") %>% 
  set_names(c("condition", "total", "dec", "oct")) %>% 
  mutate(nudge = factor(22:0))
```

Here's a look at the data.

```{r}
glimpse(d)
```

The number of persons in each condition who were vaccinated at Walmart by December 31st, 2020, is listed in the `dec` column. The total number of persons in each condition is in the `total` column. The experimental conditions are listed in the `condition` column. The `nudge` column contains the same information as `condition`, but arranged as an index variable.

If we divide `dec` by `total`, we get the proportion vaccinated in each condition. Here's a quick descriptive plot of those values.

```{r, fig.width = 8, fig.height = 3.5}
d %>% 
  mutate(p = dec / total) %>% 
  
  ggplot(aes(x = p, y = reorder(condition, p))) +
  geom_point(aes(size = total),
             alpha = 1/2) +
  scale_size_continuous(limits = c(0, NA)) +
  scale_x_continuous("% vaccinated", breaks = 29:32 / 100, labels = 29:32) +
  ylab(NULL) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid = element_blank())
```

For our statistical inference, we can model these data with a multilevel binomial model 

$$
\begin{align*}
\text{dec}_i & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \beta_0 + \zeta_{0i} \\
\zeta_{0i} & \sim \operatorname{Normal}(0, \sigma) \\
\beta_0 & \sim \operatorname{Normal}(0, 1) \\
\sigma & \sim \operatorname{Exopnential}(1),
\end{align*}
$$

where $p_i$ is the vaccine probability in each of the $i$ `nudge` conditions and and $n_i$ is the `total` number of persons within each level of `nudge`. Here's the model in **brms**.

```{r}
fit1 <- brm(
  data = d,
  family = binomial,
  dec | trials(total) ~ 1 + (1 | nudge),
  prior = c(prior(normal(0, 1), class = Intercept),
            prior(exponential(1), class = sd)),
  cores = 4,
  seed = 1,
  file = "fits/fit1.milkman2022"
)
```

Check the model.

```{r}
print(fit1)
```

It will be easiest to understand the results through a plot. Here's a version of Figure 1, as presented in the original paper.

```{r, fig.width = 8, fig.height = 3.5}
nd <- d %>% 
  select(nudge, condition, total)

fitted(fit1,
       newdata = nd,
       scale = "linear",
       summary = F) %>% 
  data.frame() %>% 
  set_names(d %>% pull(nudge)) %>% 
  mutate_all(inv_logit_scaled) %>% 
  pivot_longer(`22`:`1`) %>% 
  mutate(contrast = value - `0`) %>% 
  group_by(name) %>% 
  tidybayes::mean_qi(contrast) %>% 
  arrange(desc(as.double(name))) %>% 
  bind_cols(nd %>% filter(nudge != "0")) %>% 
  mutate(group = factor(condition, ordered = TRUE)) %>% 
  
  ggplot(aes(x = contrast)) +
  geom_pointrange(aes(xmin = .lower, xmax = .upper, y = reorder(group, contrast)),
                  fatten = 1.5) +
  geom_vline(xintercept = 0, linetype = 2, size = 1/4) +
  geom_text(data = . %>% 
              filter(name == "13") %>% 
              mutate(contrast = 0.001, 
                     label = "business as usual control: 29.7 95% CI [29.1, 30.2]"),
            aes(y = group, label = label),
            angle = 90, size = 2.5) +
  scale_x_continuous("% point increase in vaccination", breaks = 0:3 / 100, labels = 0:3) +
  labs(title = "Contrasts of all experimental conditions minus the business as usual control condition",
       y = NULL) +
  coord_cartesian(xlim = c(0.00, 0.035)) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        plot.title.position = "plot")
```

Here's a variant of that plot, which more clearly expressed the posterior for the control condition.

```{r, fig.width = 8, fig.height = 3.5}
fitted(fit1,
       newdata = nd,
       scale = "linear") %>%
  data.frame() %>% 
  mutate_all(inv_logit_scaled) %>% 
  bind_cols(nd) %>% 
  arrange(nudge) %>% 
  mutate(group = factor(condition, ordered = TRUE)) %>% 
  
  ggplot(aes(x = Estimate, xmin = Q2.5, xmax = Q97.5, y = reorder(group, Estimate))) +
  geom_pointrange(aes(color = condition == "Business As Usual Control"),
                  fatten = 1.5) +
  scale_color_manual(values = c("black", "red4"), breaks = NULL) +
  scale_x_continuous("% vaccinated", breaks = 29:32 / 100, labels = 29:32) +
  labs(title = "Flu vaccination percentages for all experimental conditions (even the business as usual control)",
       y = NULL) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        plot.title.position = "plot")
```

## Session information

```{r}
sessionInfo()
```

