---
title: "Digitize Resetar & Noell (2008)"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r, echo = F}
knitr::opts_chunk$set(fig.retina = 2.5)
options(width = 110)
```

The purpose of this file is to digitize the data from Resetar & Noell (2008; https://doi.org/10.1901/jaba.2008.41-447). We don't have their actual data file, but they displayed their data in Figure 1 (p. 450).

Load the **juicr** package.

```{r, warning = F, message = F}
library(juicr)
```

Use the `juicr::GUI_juicr()` function to digitize the data, one participant at a time, from Figure 1.

```{r, eval = F}
GUI_juicr("pics/count01.png", figureWindowSize = c(1000, 400))
GUI_juicr("pics/count02.png", figureWindowSize = c(1000, 400))
GUI_juicr("pics/count03.png", figureWindowSize = c(1000, 400))
GUI_juicr("pics/count04.png", figureWindowSize = c(1000, 400))
```

The digitized data were saved in a series of `.csv` files named `count01_juicr_extracted_points.csv` through `count04_juicr_extracted_points.csv` within the `Resetar & Noell (2008)` folder.

## Session information

```{r}
sessionInfo()
```

