---
title: "wagenmakers2016"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

The purpose of this file is to consolidate and reformat the data from Wagenmakers et al (2016; https://doi.org/10.1177/1745691616674458). The original data are openly available on the OSF at https://osf.io/9j72u/, downloadable in a zip file called `Data_FINAL.zip`. However, within that file, the data are saved in 17 separate `.csv` files, by research lab. We want them saved in a single file.

Load the **tidyverse**.

```{r, warning = F, message = F}
# library(tidyverse)

library(ggplot2)
library(tibble)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(readr)
```

Load the files and wrangle.

```{r}
# data path and individual lab file names
data_path <- "Data_FINAL/"

# column names for the .csv data
col_names <- c("subject_no", "participant_id", "condition", str_c("correct", 1:4), "correct_total", str_c("task", 1:2), str_c("rating", 1:4), "self_reported_task_performance", "comprehended_cartoons", "aware_of_the_goal", "participants_guess_of_the_studys_goal", "age", "gender", "student", "occupation_or_field_of_study")

wagenmakers2016 <- tibble(files = list.files(data_path, pattern = "*_Data.csv")) %>% 
  mutate(path = str_c(data_path, files)) %>% 
  mutate(data = map(path, ~read_csv(., 
                                    col_names = col_names, 
                                    skip = 2, 
                                    col_select = 1:22, 
                                    col_types = "dcdddddddddddddddcdddc"))) %>% 
  unnest(data) %>% 
  mutate(lab = str_extract(files, "[^_]+")) %>% 
  # update one of the lab names
  mutate(lab = ifelse(lab == "檢dogru", "Özdoğru", lab)) %>% 
  # add unique participant identifier
  mutate(sn = 1:n()) %>% 
  # rearrange columns
  select(lab, sn, everything(), -files, -path) 

# what?
glimpse(wagenmakers2016)
```

Do note, the `wagenmakers2016` data frame contains all cases, even those which are to be dropped by the exclusion criteria. Wagenmakers and colleagues had several exclusion criteria, which they detailed in the *Exclusion criteria* subsection of their *Method* (pp. 4-5). They excluded the data from participants:

* whose average cartoon rating exceeded 2.5 standard deviations from the group mean in their condition,
* who correctly guessed the goal of the study (see `aware_of_the_goal`),
* who answered "No" to the question "Did you understand the cartoons?" (see `comprehended_cartoons`), and
* who held the pen incorrectly for two or more of the four cartoons (see `correct_total`).

Here's how one might filter based on the last 3 of the 4 criteria.

```{r}
# start with n = 2612
wagenmakers2016 %>%
  # n = 2436 (drop 87 who were 1 and 89 who were NA)
  filter(aware_of_the_goal == 0) %>% 
  # n = 2190 (drop 232 who were 0, 8 who were 2, and 6 who were NA)
  filter(comprehended_cartoons == 1) %>% 
  # n = 1913 (drop 199 who were 0, 27 who were 1, 49 who were 2, and 2 who were NA)
  filter(correct_total > 2)
```

## Save

Now save the results in an external `.rda` file.

```{r}
save(wagenmakers2016, file = "wagenmakers2016.rda")
```

## Session information

```{r}
sessionInfo()
```

