---
title: "mancuso2016"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output: github_document
---

The purpose of this file is to simulate the data from Mancuso & Miltenberger (2016; https://doi.org/10.1002/jaba.267). We don't have their actual data file, they displayed their data in Figure 1 (p. 191). Here we'll work through that figure to make a data file that approximates their original data.

Load the **tidyverse** and **juicr**.

```{r, warning = F, message = F}
library(tidyverse)
library(juicr)
library(patchwork)
```

First use the `juicr::GUI_juicr()` function to digitize the data, one participant at a time, from Figure 1 in Mancuso & Miltenberger (p. 191).

```{r}
GUI_juicr("pics/rate01.png")
GUI_juicr("pics/rate02.png")
GUI_juicr("pics/rate03.png")
GUI_juicr("pics/rate04.png")
GUI_juicr("pics/rate05.png")
GUI_juicr("pics/rate06.png")
```

Now we combine the data in the .csv files, round the `session` values to integers, round the `rate` values to the third decimal place, and add participant identifiers.

```{r}
d <- tibble(files = list.files(pattern = "rate0[1-6]*_juicr_extracted_points")) %>% 
  mutate(data = map(files, read_csv)) %>% 
  unnest(data) %>% 
  transmute(sn      = str_extract(files, "\\d+") %>% as.double(),
            session = round(x.calibrated, digits = 0),
            rate    = round(y.calibrated, digits = 3)) %>% 
  arrange(sn, session) %>% 
  mutate(id = case_when(
    sn == 1 ~ "Jen",
    sn == 2 ~ "Anne",
    sn == 3 ~ "Kate",
    sn == 4 ~ "Laura",
    sn == 5 ~ "Sam",
    sn == 6 ~ "Amy"
  )) %>% 
  # convert id to an ordered factor
  mutate(id = fct_reorder(id, sn)) %>% 
  mutate(id = fct_rev(id))

# what?
head(d)
```

We need to manually add a `phase` variable, which will vary across participants and over time. I'm not aware of an elegant way to do this, so we'll just use brute force.

```{r}
d <- d %>% 
  mutate(phase = c(rep(letters[1:3], times = c(8, 5, 1)),
                   rep(letters[1:3], times = c(7, 11, 1)),
                   rep(letters[1:3], times = c(5, 6, 1)),
                   rep(letters[1:3], times = c(5, 5, 1)),
                   rep(letters[1:3], times = c(4, 7, 1)),
                   rep(letters[1:3], times = c(3, 5, 1))))
```

Here the `phase == "a"` is the baseline period, `phase == "b"` is the post-intervention assessment period, and `phase == "c"` is the follow-up.

We can now make a version of Figure 1 from the paper.

```{r}
p1 <- d %>% 
  mutate(phase = factor(phase,
                        labels = c("baseline", "postintervention", "follow-up"))) %>% 
  ggplot(aes(x = session, y = rate, color = phase, shape = phase)) +
  geom_point() +
  geom_line() +
  scale_color_manual(NULL, values = c("red3", "blue3", "green4")) +
  scale_shape_manual(values = c(16, 16, 15), breaks = NULL) +
  scale_y_continuous("rate", limits = c(0, NA), expand = expansion(mult = c(0.05, 0.25))) +
  facet_wrap(~ id, scales = "free_y", ncol = 1)

p1
```



```{r}
tibble(x = c(seq(from = 0.001, to = 0.999, by = 0.001),
             seq(from = 0.001, to = 0.999, by = 0.001))) %>% 
  mutate(phase = rep(c("baseline", "post baseline"), each = n() / 2)) %>% 
  mutate(d = ifelse(phase == "baseline",
                    dbeta(x = x, shape1 = 0.5, shape2 = 0.7),
                    dbeta(x = x, shape1 = 0.7, shape2 = 0.5))) %>% 
  mutate(seconds = x * 120 + 180) %>% 
  
  ggplot(aes(x = seconds, y = d)) +
  geom_area() +
  coord_cartesian(ylim = c(0, 8)) +
  facet_wrap(~ phase)
```

```{r}
tibble(phase = rep(c("baseline", "post baseline"), each = 1e4))%>% 
  # add the minutes variable
  mutate(minutes = ifelse(phase == "baseline", 
                          rbeta(n = n(), shape1 = 0.5, shape2 = 0.7),
                          rbeta(n = n(), shape1 = 0.7, shape2 = 0.5))) %>% 
  mutate(minutes = minutes * 2 + 3) %>% 
  
  ggplot(aes(x = minutes)) +
  geom_histogram() +
  facet_wrap(~ phase)
```

```{r}
set.seed(1)

d <- d %>% 
  # add the minutes variable
  mutate(beta = ifelse(phase == "a", 
                       rbeta(n = n(), shape1 = 0.5, shape2 = 0.7),
                       rbeta(n = n(), shape1 = 0.7, shape2 = 0.5))) %>% 
  mutate(seconds = round(beta * 120 + 180, digits = 0)) %>% 
  mutate(minutes = seconds / 60) %>% 
  # compute the counts
  mutate(count = round(rate * (seconds / 60), digits = 0))
```

Now rearrange the columns in the `d` data and save the results as `mancuso2016`.

```{r}
mancuso2016 <- d %>% 
  select(sn, id, session, phase, count, seconds, minutes, rate)

# what is this?
head(mancuso2016)
```



```{r}
p2 <- mancuso2016 %>% 
  mutate(phase = factor(phase,
                        labels = c("baseline", "postintervention", "follow-up"))) %>% 
  ggplot(aes(x = session, y = count / minutes, color = phase, shape = phase)) +
  geom_point() +
  geom_line() +
  scale_color_manual(NULL, values = c("red3", "blue3", "green4")) +
  scale_shape_manual(values = c(16, 16, 15), breaks = NULL) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0.05, 0.25))) +
  facet_wrap(~ id, scales = "free_y", ncol = 1)

# combine
p1 + p2 + plot_layout(guides = "collect")
```

```{r}
mancuso2016 %>% 
  mutate(phase = factor(phase,
                        labels = c("baseline", "postintervention", "follow-up"))) %>% 
  
  ggplot(aes(x = session, y = count, color = phase, shape = phase)) +
  geom_point(aes(size = minutes)) +
  scale_color_manual(NULL, values = c("red3", "blue3", "green4")) +
  scale_shape_manual(values = c(16, 16, 15), breaks = NULL) +
  scale_size_continuous("speech length\n(minutes)", range = c(2/3, 2), 
                        breaks = c(3, 5), limits = c(3, 5)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0.05, 0.25))) +
  facet_wrap(~ id, scales = "free_y", ncol = 1)
```



Now save the results in an external file.

```{r}
save(mancuso2016, file = "/Users/solomonkurz/Dropbox/Experimental-design-and-the-GLMM/sketches/data  simulation/mancuso2016.rda")
```

## Session information

```{r}
sessionInfo()
```


